---
title: "Download palmerpenguins data from Environmental Data Initiative"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(palmerpenguins)
```

## Download data from a URL

You can also get this penguin data directly from the Environmental Data Initiative (EDI) using the R code below. Thanks to [Julien Brun](http://brunj7.github.io/about/) for the reminder and code below to access & combine them (will get you same data as `penguins_raw`):  

```{r eval = FALSE}
# Adelie penguin data from: https://doi.org/10.6073/pasta/abc50eed9138b75f54eaada0841b9b86

uri_adelie <- "https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-pal.219.3&entityid=002f3893385f710df69eeebe893144ff"

# Gentoo penguin data from: https://doi.org/10.6073/pasta/2b1cff60f81640f182433d23e68541ce

uri_gentoo <- "https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-pal.220.3&entityid=e03b43c924f226486f2f0ab6709d2381"

# Chinstrap penguin data from: https://doi.org/10.6073/pasta/409c808f8fc9899d02401bdb04580af7
uri_chinstrap <- "https://portal.edirepository.org/nis/dataviewer?packageid=knb-lter-pal.221.2&entityid=fe853aa8f7a59aa84cdd3197619ef462"

# Combining the URIs
uris <- c(uri_adelie, uri_gentoo, uri_chinstrap)

# Downloading and importing data
library(purrr)
library(readr)
penguins_lter <- map_dfr(uris, read_csv)
```


## Compare to package data

Compared to `?penguins_raw`:

```{r}
head(penguins_raw)
```

But if that looks too messy for you, `?penguins` is simpler and tidier:

```{r}
head(penguins)
```
