---
title: "User-contributed examples"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Links to user-contributed examples, posts and vignettes: 

[**Multivariate EDA**](https://rpubs.com/friendly/penguin-biplots) by Michael Friendly

Description: These examples show the use of multivariate exploratory visualization methods -- principal component analysis and biplots -- to understand the relations among variables in the penguins data set.

[**MANOVA and HE plots**](https://rpubs.com/friendly/penguin-manova) by Michael Friendly

Description: This vignette illustrates the use of MANOVA, HE plots and canonical discriminant analysis in the analysis of the Palmer penguins data.

[**Basic data analysis with palmerpenguins**](https://www.r-bloggers.com/basic-data-analysis-with-palmerpenguins/) by jhk0530

Description: Examples of EDA and statistical analysis with penguins

### To add your examples: 

Please submit a GitHub [issue](https://github.com/allisonhorst/palmerpenguins/issues) that includes a short title, link, author name(s), and brief (1 - 2 sentence) description.
